---
title: "An Analysis of Covid Data"
author: "Taylor Edwards"
date: "2023-05-30"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```
## Introduction

The analysis presented in this document focuses on exploring and understanding a dataset provided by Johns Hopkins University. The dataset contains detailed information on COVID19 cases and deaths throughout the world. The objective is to simply explore the data to see if any points of interest are uncovered.


## Housekeeping

To begin the analysis, it is necessary to install and load certain packages that will be used throughout the document. These packages provide various functionalities for data manipulation, visualization, mapping, and reporting. Please make sure to install the packages listed below.

```{r get_packages, include = TRUE, message=FALSE}
library(tidyverse)
library(lubridate)
```

## The Data

After reading in and manipulating the data into a usable form, we end up with both a global and a US-only covid data set.

```{r get_data, echo = FALSE}
urls_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
filenames <- c("time_series_covid19_confirmed_global.csv",
               "time_series_covid19_deaths_global.csv",
               "time_series_covid19_confirmed_US.csv",
               "time_series_covid19_deaths_US.csv",
               "time_series_covid19_recovered_global.csv")
urls <- str_c(urls_in,filenames)
```

```{r import_data, echo = FALSE, message = FALSE}
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
US_cases <- read_csv(urls[3])
US_deaths <- read_csv(urls[4])
global_recovered <- read_csv(urls[5])
global_population <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv")
```

```{r tidy_global_data, echo = FALSE}
global_cases_tidy <- global_cases %>%
  pivot_longer(cols = -c('Province/State',
                         'Country/Region', Lat, Long),
               names_to = "date",
               values_to = "cases") %>% select(-c(Lat,Long))

global_deaths_tidy <- global_deaths %>%
  pivot_longer(cols = -c('Province/State',
                         'Country/Region', Lat, Long),
               names_to = "date",
               values_to = "deaths") %>% select(-c(Lat,Long))
```

```{r combine_global_data, echo = FALSE}
global <- global_cases_tidy %>%
  full_join(global_deaths_tidy) %>%
  rename(Country_Region = 'Country/Region',
         Province_State = 'Province/State') %>%
  mutate(date = mdy(date))

global <- global %>% filter(cases > 0)

global <- global %>% 
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)

global[is.na(global)] <- ""

# global_population <- global_population %>% 
#   unite("Combined_Key",
#         c(Province_State, Country_Region),
#         sep = ", ",
#         na.rm = TRUE,
#         remove = FALSE)

global_population_temp <- global_population %>% 
  select("Combined_Key", "Population")



global_tidy <- global %>% 
  left_join(global_population_temp, by = c("Combined_Key")) %>%
  select(Combined_Key, Country_Region, Province_State,  
        date, cases, deaths,
        Population)

```

```{r tidy_US_data, echo = FALSE}
US_cases_tidy <- US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat,Long_))

US_deaths_tidy <- US_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat,Long_))

US <- US_cases_tidy %>% full_join(US_deaths_tidy)

```
Here's the global covid data:

```{r show_data, include = TRUE}
head(global_tidy)
```

Here's the US-only covid data:

```{r show_US_data, include = TRUE}
head(US)
```

## Analysis

We have decided to focus on US covid data only. First, we work to summarize the data and add some fields that add some context to the data. 

```{r US_by_state}
US_by_state <- US %>% 
  group_by(Province_State, Country_Region, date) %>% 
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>% 
  mutate(deaths_per_mil = deaths*1000000/Population) %>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mil, Population) %>% 
  ungroup()
```

```{r US_totals}
US_totals <- US_by_state %>% 
  group_by(Country_Region, date) %>% 
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>% 
  mutate(deaths_per_mil = deaths*1000000/Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mil, Population) %>% 
  ungroup()
```

The following figure shows the cumulative total cases and deaths for the entire country:


```{r viz_cases_deaths_total}
US_totals %>% 
  filter(cases > 0) %>% 
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US", y = NULL, color = "")
```


The previous figure could be a little misleading. Perhaps, we are more interested in how new cases and deaths look over time. The following figure shows the cumulative total new cases and new deaths for the entire country:

```{r new_cases_deaths}
US_totals <- US_totals %>% 
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

US_totals %>% 
  filter(cases > 0) %>% 
  ggplot(aes(x = date, y = new_cases)) +
  #geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = deaths, color = "new_deaths")) +
  geom_point(aes(y = deaths, color = "new_deaths")) +
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US", y = NULL, color = "")
```


Now let's look at a single state, California. The following figures show how covid cases and deaths have changed over time. Similar the the above figures, we also show how new cases and new deaths have changed with time in California.

```{r state_CD_CA, warning= FALSE}
state <- "California"

US_by_state <- US_by_state %>% 
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

US_by_state %>% 
  filter(Province_State == state) %>% 
  filter(cases > 0) %>% 
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", state), y = NULL, color = "")



US_by_state %>% 
  filter(Province_State == state) %>% 
  filter(cases > 0) %>% 
  ggplot(aes(x = date, y = new_cases)) +
  #geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  #geom_line(aes(y = deaths, color = "new_deaths")) +
  geom_point(aes(y = deaths, color = "new_deaths")) +
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("New cases and deaths in ", state), y = NULL, color = "")

US_state_totals <- US_by_state %>% 
  group_by(Province_State) %>% 
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population),
            cases_per_thou = 1000 * cases / Population,
            deaths_per_thou = 1000 * deaths / population) %>% 
  filter(cases > 0, population >0)

US_state_totals %>% slice_min(deaths_per_thou, n = 10)
```

Now let's do some comparison between two states to see if we notice anything interesting. Below is a figure comparing covid data for California and Texas. We can see that the death rate in Texas appears to be greater than California's over time. Texas' death rate was lower during the beginning of the pandemic. One could assume that this is because of Texes' lower population density. The death rate in Texas may have overtaken California's due to more relaxed covid safety protocols.

```{r state_deaths_CA_v_Texas, echo = FALSE}
US_by_state %>% 
  filter(Province_State == c("California","Texas") ) %>% 
  filter(cases > 0) %>% 
  ggplot(aes(x = date, y = deaths_per_mil, group = Province_State)) +
    geom_point(aes(color = Province_State)) +
    theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "Deaths in CA vs TX", y = "Deaths/million", color = "")


```

### Modeling

In the figure 'COVID19 in US', we observed a spike in cases and deaths at the beginning of the pandemic. As time went on, cases and deaths level off to a less drastic rate. During this time, more covid protocols were put in place with varying degrees of strictness. What the following models attempt to answer is:

-   Is the relationship between cases and deaths linear?
-   Does this relationship change for different stages of the pandemic?

```{r model_data, echo = FALSE}
#Model the cases vs deaths data for 2020 only
#Get US totals 2in 2020
US_total_2020 <- US_by_state %>%
  filter(year(date) == 2020) %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
  mutate(deaths_per_thou = deaths * 1000 / Population) %>%
  mutate(cases_per_thou = cases *1000 / Population) %>% 
  select(Country_Region, date, cases, deaths, cases_per_thou, deaths_per_thou, Population) %>%
  ungroup()

#Create linear model of cases vs deaths
mod1 <- lm(deaths_per_thou ~ cases_per_thou, data = US_total_2020)
summary(mod1)

slope1 <- mod1$coefficients["cases_per_thou"]
intercept1 <- mod1$coefficients["(Intercept)"]

pred_y <- slope1 * US_total_2020$cases_per_thou + intercept1
  
US_total_2020 %>% 
  ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou),color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred_y), color = "red") +
  geom_text(x = 20, y = 1, label = paste0("R-squared = ", round(summary(mod1)$r.squared, 3)))  +
  labs(title = "Cases vs Deaths 2020", x = "Cases/Thousand", y = "Deaths/Thousand", color = "")



#Now complete similar model for the entire pandemic timeframe
US_total <- US_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
  mutate(deaths_per_thou = deaths * 1000 / Population) %>%
  mutate(cases_per_thou = cases *1000 / Population) %>% 
  select(Country_Region, date, cases, deaths, cases_per_thou, deaths_per_thou, Population) %>%
  ungroup()

#Create linear model of cases vs deaths
mod2 <- lm(deaths_per_thou ~ cases_per_thou, data = US_total)
summary(mod2)

slope2 <- mod2$coefficients["cases_per_thou"]
intercept2 <- mod2$coefficients["(Intercept)"]

pred_y <- slope2 * US_total$cases_per_thou + intercept2
  
US_total %>% 
  ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou),color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred_y), color = "red") +
  geom_text(x = 100, y = 3, label = paste0("R-squared = ", round(summary(mod2)$r.squared, 3)))+
  labs(title = "Cases vs Deaths after 2020", x = "Cases/Thousand", y = "Deaths/Thousand", color = "")
```

The models show that cases are linearly related to deaths for both 2020 and all years after 2020. The models also show that the linear relationship between the two variables is actually stronger after 2020. The contributing factors to this phenomena are likely numerous and hard to measure. 

In further analysis, one could model the relationship between new cases and new deaths for these same time periods. 

### Bias & Limitations

It is important to acknowledge any biases or limitations in the data and analysis. First, we assume that the data was gathered in a fair, unbiased way. Second we acknowledge that covid can be a highly polarizing topic. Choosing what data to analyze and what trends to show can be indicators of bias. The reader should know that bias may affect the analysis within the report as stringent fairness techniques were not implemented.